---
title: "simulateOccJSDMData"
output: 
  rmarkdown::html_vignette:
    fig.width: 7
    fig.height: 4.5
    warning: false
    message: false
vignette: >
  %\VignetteIndexEntry{simulateOccJSDMData}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warnings = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  comment = "#>"
)

options(warning = -1)
```

```{r setup, include = FALSE}
library(occJSDM)
library(ggplot2)
library(dplyr)
library(tibble)
```

`simulateOccJSDMData()` simulates a full occJSDM-style dataset: species
occupancy driven by environmental covariates, species traits, and
(optionally) a spatially autocorrelated random field, plus a two-stage
eDNA collection/detection process on top of the true occupancy states.

The function takes three lists of settings, plus a `useSpatField`
switch:

- `list_datasettings`: sample-size and dimension settings -- `n`
  (sites), `S` (species), `g` (number of observed trait covariates), `M`
  (visits per site), `P` (primers), `K` (PCR replicates per
  site-visit-primer), `ncov_psi` (occupancy covariates), and
  `ncov_theta` (eDNA collection covariates).
- `list_jsdmParams`: JSDM structural parameters -- `gt` (latent trait
  dimensions), `d` (latent factor dimensions capturing residual species
  co-occurrence), `ds` (spatial latent factor dimensions),
  `sigma_b`/`sigma_bs`/`sigma_ts`/`sigma_h` (variance components), and
  `l_s` (spatial length scale).
- `list_params`: detection-process parameters -- `p` and `q` (detection
  and false-positive amplification rates, `P` x `S` matrices), `theta0`
  and `theta_baseline` (baseline eDNA collection probabilities given
  absence/presence).
- `useSpatField` (an element of `list_jsdmParams`, with no default -- it
  must be supplied): site coordinates (`Xs`) are simulated identically
  either way (independent `Uniform(0, 1)` draws per site) -- what
  differs is whether they're *used*. If `TRUE`, occupancy incorporates a
  spatially autocorrelated Gaussian process built from `Xs` via a
  squared-exponential kernel. If `FALSE`, that spatial term is set to an
  exact zero matrix instead, so `Xs` is simulated and returned but has
  no influence on occupancy.

## Setting up shared simulation parameters

We use the same dimension, trait, and detection-process settings for
both datasets, so that the *only* difference between them is whether the
spatial field is switched on.

```{r}
set.seed(5162)

n <- 100          # sites
M <- rep(3, n)    # visits per site
P <- 3            # primers/markers
S <- 10           # species/OTUs
g <- 3            # observed trait covariates
gt <- 2           # unobserved (latent) trait dimensions
d <- 2            # latent factor dimensions (residual co-occurrence)
ds <- 2           # spatial latent factor dimensions
ncov_psi <- 2     # environmental covariates for occupancy
ncov_theta <- 2   # covariates for eDNA collection probability

N <- sum(M)
K <- rep(2, P * N)  # PCR replicates per site-visit-primer combo

list_datasettings <- list(
  n = n, S = S, g = g, M = M, P = P, K = K,
  ncov_psi = ncov_psi, ncov_theta = ncov_theta
  )

list_jsdmParams <- list(
  gt = gt,
  # gt is the number of *unobserved* (latent) trait dimensions, on top of the
  # g observed traits. These let species' environmental responses be partly
  # explained by traits the study never measured.
  d = d,
  # d is the number of latent factor dimensions. Factor scores U (n x d) and
  # loadings L (d x S) together generate residual co-occurrence -- correlation
  # between species left over once the environmental covariates are accounted
  # for. At d = 0 species would be conditionally independent.
  ds = ds,
  # ds is the rank of the *cross-species* spatial correlation, used only when
  # useSpatField = TRUE. At ds > 0 each species gets its own spatial field,
  # correlated across species at rank ds. At ds = 0 a single field is drawn
  # and shared identically by every species.
  sigma_b = 0.5,
  # sigma_b is the SD of the residual environmental coefficients: the part of
  # each species' response to the ncov_psi covariates that its traits do not
  # explain (drawn Normal(0, sigma_b^2)). Larger values make species respond
  # more idiosyncratically relative to their traits.
  sigma_bs = 0.5,
  # sigma_bs is the equivalent residual SD for the *spatial* coefficients.
  # Note it has no effect on the simulated data: the simulator sets the
  # residual spatial term to an exact zero matrix, so nothing is ever drawn
  # with this SD. It is carried in the returned true parameters and read by
  # the fitting side only. Supplied here for completeness.
  sigma_ts = 0.5,
  # sigma_ts is accepted by the simulator and never referenced in its body,
  # so it affects nothing at all. Supplied only because the argument exists.
  sigma_h = 1,
  # sigma_h is the SD of the latent factor scores (U drawn Normal(0,
  # sigma_h^2), iid across sites and factors). With the loadings L fixed by
  # construction, this is what sets the strength of the residual
  # co-occurrence signal.
  sigma_s = .5,
  # sigma_s is the amplitude of the spatial Gaussian process. The kernel is
  # sigma_s * exp(-dist^2 / (2 * l_s^2)), which equals sigma_s at zero
  # distance, so this is the *marginal variance* of the spatial field rather
  # than its SD: at 0.5 a site's field value has SD sqrt(0.5), about 0.71.
  # Used only when useSpatField = TRUE.
  l_s = 0.3,
  # l_s is the length scale of that same kernel, in the units of the site
  # coordinates. Xs is drawn Uniform(0, 1) on each axis, so the domain is the
  # unit square and l_s = 0.3 correlates occupancy over roughly a third of
  # it -- wide enough to give visible spatial structure at n = 100 sites,
  # without making the field so smooth it is effectively constant.
  useSpatField = TRUE
  # useSpatField switches the spatial term on. As noted above, Xs is
  # simulated either way; at FALSE the field is replaced by an exact zero
  # matrix and has no influence on occupancy.
  )

# add splinevars to list_params

list_params <- list(
  p = matrix(runif(P * S, 0.3, 0.6), P, S),   
  # p is the Stage 2 (PCR/lab): true-positive detection rate per PCR 
  # replicate, given the species was collected into the sample (w = 1). 
  # Set at the range ~0.3-0.6 so that a single PCR replicate
  # often misses a truly-collected species -- this makes
  # the number of species detected increase visibly with K (replicates
  # per sample), rather than saturating at K = 1.
  q = matrix(runif(P * S, 0.01, 0.05), P, S),  
  # q is the Stage 2 (PCR/lab): false-positive rate per PCR replicate, given 
  # the species was *not* collected (w = 0). Here it is set low to simulate
  # few false positives.
  theta0 = runif(S, 0.02, 0.1),                
  # theta0 is the Stage 1 (field collection): false-positive collection rate 
  # given the species is truly absent (z = 0), e.g. contamination. 
  # Kept low to simulate low contamination.
  theta_baseline = runif(S, 0.15, 0.4)          
  # theta_baseline is the Stage 1 (field collection): baseline collection 
  # probability given the species is truly present (z = 1). Set at the
  # range ~0.15-0.4 to simulate the scenario where present species are often 
  # missed by any single field sample.
  ) 

  # Together, low p and low theta_baseline mean that true detections are low
  # in both stages 1 and 2. 
```

## Simulating a dataset for occJSDM()

We fix a seed immediately before this call (and reuse it below for
`sim_nospat`) so that both simulations draw the *same* site coordinates
`Xs` -- `Xs` is generated early in `simulateData()`, before the code
path diverges on `useSpatField`, so reseeding to the same value before
each call keeps `Xs` (and `X_psi`, `Tr`) identical across the two
datasets. This isolates `useSpatField`'s effect for a cleaner comparison
later.

```{r}
sim_seed <- 7391

set.seed(sim_seed)
sim <- simulateOccJSDMData(
  list_datasettings, 
  list_params, 
  list_jsdmParams,
  model = "two_stage"
  )

data <- sim$data_list # data for runOccJSDM()

str(data, max.level = 1)
head(data$info); dim(data$info); names(data$info) # study design columns (Site, Sample, Primer), occupancy covariates (X_psi.EnvCov.n), spatial covariates (Xs.n), and collection covariates (X_theta.n)
head(data$OTU); dim(data$OTU) # rows = PCR replicates, columns = OTUs
head(data$traits); dim(data$traits) # species x traits


# save as default dataset for occJSDM package
# sampledata <- data
# usethis::use_data(sampledata, overwrite = TRUE)
```

## Spatial autocorrelation

### Simulating without spatial autocorrelation

```{r}
list_jsdmParams_nospat <- modifyList(list_jsdmParams, 
                                     list(useSpatField = FALSE)
                                     )

set.seed(sim_seed)
sim_nospat <- simulateOccJSDMData(
  list_datasettings, list_params, list_jsdmParams_nospat,
  model = "two_stage"
  )

str(sim_nospat$data_list, max.level = 1)
```

`data_list` always contains the trait matrix `traits` (`S` x `g`) and
site coordinates (`info$Xs.1`/`info$Xs.2`, one row per PCR replicate,
repeated within a site), regardless of `useSpatField` -- site
coordinates are simulated either way, but only feed into occupancy when
`useSpatField = TRUE`.

### Simulating with spatial autocorrelation

We rename `sim` to `sim_spat`, to remind ourselves that this dataset has
spatial factors. The spatial component uses `ds` latent spatial factors
and length scale `l_s` to build a Gaussian-process-correlated random
field over `Xs`, which is added to each species' linear predictor for
occupancy.

```{r}
sim_spat <- sim

str(sim_spat$data_list, max.level = 1)
```

### Checking how much variance is spatial

Each simulation's `true_params$jsdmParams_true$varPart` decomposes the
linear predictor's variance per species into Environmental, Spatial, and
Biotic components (plus a Total row of residual/intercept variance).
This is a quick way to check that turning on `useSpatField` actually
contributes spatial variance.

```{r}
sim_nospat$true_params$jsdmParams_true$varPart
sim_spat$true_params$jsdmParams_true$varPart
```

As expected, the "Spatial" column is (numerically) negligible when
`useSpatField = FALSE`, and becomes a non-trivial share of variance --
though it varies considerably across species -- when
`useSpatField = TRUE`.

### Visualizing simulated occupancy across sites

We can compare the true occupancy states (`z_true`) for a single
species, plotted at its site coordinates, with and without the spatial
field switched on.

```{r}
species_idx <- 5

# One row per site: info repeats Site/Xs across PCR replicates, so
# de-duplicate before joining to the site-level z_true matrix.
site_coords <- function(sim) {
  info <- sim$data_list$info
  info[!duplicated(info$Site), c("Site", "Xs.1", "Xs.2")]
}

df_compare <- bind_rows(
  tibble(
    x = site_coords(sim_nospat)$Xs.1,
    y = site_coords(sim_nospat)$Xs.2,
    occupied = factor(sim_nospat$true_params$z_true[, species_idx], levels = c(0, 1)),
    field = "useSpatField = FALSE"
  ),
  tibble(
    x = site_coords(sim_spat)$Xs.1,
    y = site_coords(sim_spat)$Xs.2,
    occupied = factor(sim_spat$true_params$z_true[, species_idx], levels = c(0, 1)),
    field = "useSpatField = TRUE"
  )
)

ggplot(df_compare, aes(x, y, color = occupied)) +
  geom_point(size = 2.5) +
  facet_wrap(~field) +
  labs(
    title = paste0("Simulated true occupancy for species ", species_idx),
    x = "Spatial coordinate 1", y = "Spatial coordinate 2", color = "Occupied (z)"
  )
```

Because occupancy is thresholded to 0/1 and the environmental and biotic
terms often explain more variance than the spatial term (see the
`varPart` tables above), spatial clustering is not always obvious by eye
in the binary occupancy map, even when the underlying field is genuinely
spatially structured. The continuous linear predictor
(`true_params$jsdmParams_true$eta`) is a less noisy way to inspect the
spatial pattern directly:

```{r}
df_eta <- tibble(
  x = site_coords(sim_spat)$Xs.1,
  y = site_coords(sim_spat)$Xs.2,
  eta = sim_spat$true_params$jsdmParams_true$eta[, species_idx]
)

ggplot(df_eta, aes(x, y, color = eta)) +
  geom_point(size = 3) +
  scale_color_viridis_c() +
  labs(
    title = paste0("Linear predictor (eta) for species ", species_idx, ", useSpatField = TRUE"),
    x = "Spatial coordinate 1", y = "Spatial coordinate 2", color = "eta"
  )
```

To make the spatial signal visually dominant, increase `l_s` (broader,
smoother spatial patches) and/or reduce `sigma_h`/`sigma_b` (less
non-spatial noise from latent factors and environmental covariates).

## How traits shape species-specific occupancy responses

Species' responses to environmental covariates are not simulated
independently: each species' coefficient vector for `X_psi` is built
partly from its trait values `Tr`, via a trait-covariate coefficient
matrix `G` (`g` x `ncov_psi`). Specifically,

$$B = (Tr \, G + A \, C + B_t)^\top,$$

where `A` and `C` represent an *unobserved* trait dimension and its
response, and `Bt` is residual, trait-independent variation. `B`
(`ncov_psi` x `S`) is the matrix of true environmental-covariate
coefficients actually used to simulate occupancy.

```{r}
G_true <- sim_nospat$true_params$jsdmParams_true$G
B_true <- sim_nospat$true_params$jsdmParams_true$B
Tr_true <- sim_nospat$data_list$traits

G_true   # g x ncov_psi: how each trait shifts each covariate's coefficient
```

`G`'s entries are drawn from `{-1, 0, 1}`, so each trait either has no
effect, a positive effect, or a negative effect on a given covariate's
coefficient. We can isolate the trait-driven component of `B`,
`Tr %*% G`, and compare it to the full coefficient matrix to see how
much of species-to-species variation in the covariate response is
attributable to traits versus the unobserved/residual terms.

```{r}
trait_component <- Tr_true %*% G_true   # S x ncov_psi

df_trait <- tibble(
  species = paste0("OTU_", seq_len(S)),
  trait2 = Tr_true[, 2],
  coef_cov1_full = B_true[1, ],
  coef_cov1_trait_only = trait_component[, 1]
)

df_trait
```

Plotting each species' true coefficient for the first covariate against
its value for the second trait shows the trait-driven relationship,
though the full coefficient (which also includes the unobserved-trait
and residual terms) will scatter around it rather than fall exactly on a
line.

```{r}
ggplot(df_trait, aes(trait2, coef_cov1_full)) +
  geom_point(size = 2.5) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed") +
  labs(
    title = "Species' occupancy response to covariate 1 vs. trait 2",
    x = "Trait 2 value", y = "True coefficient for covariate 1 (B)"
  )
```

The sign and magnitude of the slope is governed by `G_true[2, 1]` (trait
2's effect on covariate 1) -- shown above to be nonzero, so a
relationship is expected here, with its sign matching `G_true[2, 1]`,
attenuated by the unobserved-trait pathway (`A %*% C`) and residual
noise (`Bt`). Had `G_true[2, 1]` been `0` instead, species' coefficients
would vary with trait 2 only through those other pathways, and no
systematic relationship would be expected.
