---
title: "vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE, }
library(occPlus)
```

The model can be run using the `runOccPlus` function.

The arguments of the function are:

- data: which we explain below;
- d: number of factors;
- occCovariates: vector of the name of the covariates for the occupancy probabilities. 
- ordCovariates: vector of the name of the covariates for the ordination. 
- detCovariates: vector of the name of the covariates for the detection probabilities. 


The names of the covariates in `occCovariates`, `ordCovariates` and `detCovariates`
should match the column name in data$info.

The data object is a list of two elements: `info` and `OTU`. 

`data$info` is a data.frame with $N$ rows, where $N$ is the total
number of replicates analyzed and columns:

- `Site`: name (or index) of the site in which this replicate was collected;
- `Sample`: name (or index) of the sample from which this replicate was analyzed;
- `Primer`: name (or index) of the primer used in this replicate;


```{r}
head(data$info)
```

`data$OTU` is a matrix of dimension $N \times S$, where $S$ is the number of 
species, where each element is the number of reads of the species in that replicate.
If data are missing, these can be coded as `NA`.


```{r}
head(data$OTU)
```
