---
title: "occJSDM: occupancy maps, truth against prediction"
output:
  html_document:
    self_contained: true
    theme: null
    css: occupancy_maps.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      dev = "png", dpi = 170)
library(ggplot2)
# Paths resolve from the package root, wherever this is rendered from, so the
# document works from a clean checkout. knitr sets the working directory to
# this file's own directory, hence the walk up.
root <- local({
  d <- normalizePath(".", mustWork = FALSE)
  for (i in 1:6) {
    if (file.exists(file.path(d, "DESCRIPTION"))) break
    if (identical(dirname(d), d)) break
    d <- dirname(d)
  }
  d
})
map_rds <- file.path(root, "dev/simstudy/mapdata.rds")
if (!file.exists(map_rds))
  stop("dev/simstudy/mapdata.rds not found. Run first:\n",
       "  RCPP_PARALLEL_NUM_THREADS=1 Rscript dev/simstudy/make_occupancy_maps.R")
m <- readRDS(map_rds)
ps <- m$per_species
long <- m$long
# Newest occupancy summary from whatever study run is on disk.
csv <- local({
  f <- list.files(file.path(root, "dev/simstudy/results"),
                  pattern = "^simstudy-occupancy-.*\\.csv$", full.names = TRUE)
  if (length(f) == 0)
    stop("no simstudy-occupancy-*.csv in dev/simstudy/results. ",
         "Run dev/simstudy/run_study.R first.")
  read.csv(f[order(file.mtime(f), decreasing = TRUE)][1],
           stringsAsFactors = FALSE)
})

theme_map <- theme_minimal(base_size = 11) +
  theme(panel.grid = element_blank(),
        axis.text = element_blank(), axis.title = element_blank(),
        axis.ticks = element_blank(),
        strip.text = element_text(face = "bold", size = 10),
        legend.position = "bottom",
        legend.key.width = unit(2.2, "cm"),
        legend.key.height = unit(0.35, "cm"))
```

## What you are looking at

One simulated dataset: **100 sites, 10 species**, the `base` scenario, replicate 1 of the R = 100 study. Every site has a real 2-D coordinate, so occupancy can be drawn as an actual map.

The left column is the **truth** we simulated. The right column is what `runOccJSDM()` recovered, knowing none of it. Rings mark sites where the species genuinely is present.

```{r maps, fig.width=9, fig.height=11.5}
sel <- ps$species[order(-ps$psi_cor)][c(1, 4, 7, 10)]
lab <- setNames(
  sprintf("species %d  |  present at %.0f%% of sites  |  psi_cor %.2f",
          ps$species, 100 * ps$occ_true, ps$psi_cor), ps$species)

d <- long[long$species %in% sel, ]
d$panel <- factor(lab[as.character(d$species)], levels = lab[as.character(sel)])

dl <- rbind(
  transform(d[, c("panel","x","y","z_true")], value = d$psi_true,
            what = "TRUTH: simulated occupancy probability"),
  transform(d[, c("panel","x","y","z_true")], value = d$psi_est,
            what = "MODEL: estimated occupancy probability"))
dl$what <- factor(dl$what, levels = c("TRUTH: simulated occupancy probability",
                                      "MODEL: estimated occupancy probability"))

ggplot(dl, aes(x, y)) +
  geom_point(aes(colour = value), size = 3.1) +
  geom_point(data = dl[dl$z_true == 1, ], shape = 21, size = 4.4,
             colour = "grey15", stroke = 0.35, fill = NA) +
  facet_grid(panel ~ what, switch = "y") +
  scale_colour_viridis_c(name = "probability of occupancy",
                         limits = c(0, 1), option = "D") +
  coord_equal() +
  theme_map +
  theme(strip.text.y.left = element_text(angle = 0, hjust = 0, size = 9))
```

Reading these: the model is not trying to reproduce the rings, it is trying to reproduce the left-hand surface. A species is well recovered when the right map has its bright and dark regions in the same places as the left, whatever the overall brightness.

The top row is close to the best this model does. The bottom row is close to the worst, and the failure is visible as a right-hand map that is comparatively flat: the model has spread the species roughly evenly instead of concentrating it where it belongs.

## Every species at once

```{r scatter, fig.width=9, fig.height=4.6}
long$panel <- factor(sprintf("sp %d  (r = %.2f)", long$species,
                             ps$psi_cor[match(long$species, ps$species)]),
  levels = sprintf("sp %d  (r = %.2f)", ps$species[order(-ps$psi_cor)],
                   ps$psi_cor[order(-ps$psi_cor)]))

ggplot(long, aes(psi_true, psi_est)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60",
              linetype = "dashed", linewidth = 0.4) +
  geom_point(aes(colour = factor(z_true)), size = 0.85, alpha = 0.8) +
  scale_colour_manual(values = c("0" = "grey65", "1" = "#2B7A78"),
                      labels = c("truly absent", "truly present"),
                      name = NULL) +
  facet_wrap(~ panel, nrow = 2) +
  labs(x = "true occupancy probability", y = "estimated") +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom",
        strip.text = element_text(face = "bold", size = 8.5))
```

Each point is one site. The dashed line is perfect recovery. Points hugging it mean the model has the site-by-site pattern right; a horizontal smear means it has settled on roughly one probability everywhere and is telling you nothing about location.

Note that the estimates are compressed toward the middle in every panel: the model rarely commits to a probability near 0 or 1. That is shrinkage, and it is why `psi_cor` is the statistic here rather than agreement on the raw values.

## Across all ten scenarios

This is the part that comes from the CSV: 1000 species-fits per scenario, so each point has 100 datasets behind it rather than one.

```{r scenarios, fig.width=9, fig.height=4.2}
csv$scenario <- factor(csv$scenario, levels = csv$scenario[order(csv$psi_cor_mean)])
ggplot(csv, aes(y = scenario)) +
  geom_vline(xintercept = 0, colour = "grey75", linewidth = 0.4) +
  geom_errorbarh(aes(xmin = psi_cor_q05, xmax = psi_cor_q95), height = 0,
                 colour = "grey60", linewidth = 0.6) +
  geom_point(aes(x = psi_cor_mean), size = 2.6, colour = "#1F5F6B") +
  geom_point(aes(x = psi_cor_q50), size = 1.5, shape = 21, fill = "white",
             colour = "#1F5F6B") +
  labs(x = "psi_cor: how well occupancy is placed (filled = mean, hollow = median, bar = middle 90% of species-fits)",
       y = NULL) +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        axis.title.x = element_text(size = 8.5))
```

Nine of the ten scenarios sit together around 0.58 to 0.67. **`low_information` is the outlier and it is not close**: mean 0.15, and its lower tail runs below zero, meaning a fair number of species-fits place the species *worse than a coin flip*. That cell is 40 sites with weak detection, and it is the configuration to check your own survey against before trusting a map like the ones above.

The wide bars everywhere else are the honest part of this picture. Even in the healthy cells, individual species-fits range from near zero to above 0.9. A single fit tells you much less than its own tidy-looking numbers suggest, which is the whole reason the study fits each scenario a hundred times.
